---
title: "NDBC Historical Data"
author: "Brian High"
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output:
  html_document:
    keep_md: yes
  pdf_document: default
---

## Clear Workspace

```{r clear_workspace, echo=TRUE, message=FALSE, warning=FALSE}
# Clear the workspace, unless you are running in knitr context
if (!isTRUE(getOption('knitr.in.progress'))) {
    closeAllConnections()
    rm(list = ls())
}
```

## Configure Settings

Configure settings for the rendering of this document.

```{r set_knitr_options, echo=TRUE, message=FALSE}
suppressMessages(library(knitr))
opts_chunk$set(tidy=FALSE, cache=FALSE, echo=TRUE, message=FALSE)
```

Configure the NDBC Station ID, Station Name, local timezone and number of days 
of weather observations to plot. 

```{r}
stn.id    <- '46041'            # ID List: http://www.ndbc.noaa.gov/to_station.shtml
stn.name  <- 'Cape Elizabeth'   # https://www.google.com/search?q=ndbc+neah+bay
tz.string <- 'US/Pacific'       # Convert NDBC date/times from UTC to this timezone
days      <- 3                  # Number of days to plot
```

## Retrieve NDBC Data

Download the text data file for NOAA NDBC Station ID `r stn.id` (`r stn.name`).

```{r}
cols <- c('YY', 'MM', 'DD', 'hh', 'mm', 'WDIR', 'WSPD', 'GST', 'WVHT', 'DPD', 
          'APD', 'MWD', 'PRES', 'ATMP', 'WTMP', 'DEWP', 'VIS', 'PTDY', 'TIDE')
site <- 'http://www.ndbc.noaa.gov/data/realtime2/'
ndbc <- read.fwf(
    file=url(paste(site, stn.id, '.txt', sep='')), col.names=cols, skip=2,
    widths=c(4, 3, 3, 3, 3, 4, 5, 5, 6, 6, 6, 4, 7, 6, 6, 6, 5, 5, 6), 
    na.strings=c('MM'), stringsAsFactors=FALSE, strip.white=TRUE)
```

## Calculate Date Range and Subset Data

Convert dates to POSIX date format and select only last `r days` days of wind 
gusts and wave height.

```{r}
# Determine start and end dates and convert dates to strings for use in plot
end.date <- as.POSIXct(Sys.time(), tz=tz.string)
start.date <- end.date - as.difftime(days, units = 'days')
end.date.string <- format(end.date, format='%B %d, %Y')
start.date.string <- format(start.date, format='%B %d, %Y')

# Convert date components of raw data to a date-time format for calculations
ndbc$DATE <- as.POSIXct(paste(ndbc$YY, ndbc$MM, ndbc$DD, ndbc$hh, ndbc$mm, 
                              sep='.'), format = '%Y.%m.%d.%H.%M', tz='UTC')
attr(ndbc$DATE, 'tzone') <- tz.string
ndbcww <- na.exclude(ndbc[ndbc$DATE > start.date, c('DATE', 'WVHT', 'GST')])
```

## Plot Wind Gusts and Wave Height

Plot wind gusts and wave height by date using base R graphics.

```{r, echo=FALSE, eval=FALSE}
# Save the plot as a PNG image named after the station ID and current date/time
plot.date <- format(end.date, format='%Y-%m-%d_%H%M%S')
plot.file <- paste(stn.id, '_', plot.date, '.png', sep='')
png(plot.file, width=640, height=480, units='px')
```

```{r}
# Set graphical parameters (set margins)
par(mar=c(5, 4, 4, 5) + .1)

# Plot wind gusts with conversion from meters per second (m/s) to knots (kt)
plot(ndbcww$DATE, ndbcww$GST*1.943844, 
     type='l', col='red', ylab='Wind Gusts (kt)', xlab='')

# Set graphical parameters (plot next plot onto the previous plot frame)
par(new=T)

# Plot wave height, with conversion from meters (m) to feet (ft)
plot(ndbcww$DATE, ndbcww$WVHT*3.28084, 
     type='l', col='blue', xaxt='n', yaxt='n', xlab='', ylab='')

# Add additional plot settings (axis labels, title, legend, and footer)
axis(4)
mtext('Wave Height (ft)', side=4, line=3)
title(main=paste('Wind Gusts and Wave Height, NDBC Station', 
                 stn.id, '-', stn.name))
legend('topleft', legend=c('Wind Gusts', 'Wave Height'), 
       col=c('red', 'blue'), lty=1, cex=0.75)
mtext(paste(start.date.string, '-', end.date.string), side=1, line=3)
mtext('Source: National Data Buoy Center, NOAA', font=3, side=1, line=4)
```

```{r, echo=FALSE, eval=FALSE}
# Close the graphics device
dev.off()
```

## Data Sources

Plot data are from [U.S. Dept. of Commerce](https://www.commerce.gov/) 
[National Oceanic and Atmospheric Administration](http://www.noaa.gov/) (NOAA) 
[National Weather Service](http://www.weather.gov/) (NWS) 
[National Data Buoy Center](http://www.ndbc.noaa.gov/) (NDBC).

## License

The RMarkdown source document and all rendered reports generated from it are 
licensed under the [Creative Commons CC0 1.0 Universal Public Domain Dedication](https://creativecommons.org/publicdomain/zero/1.0/).
