---
title: "R Bootcamp: Pre-Class Survey"
author: "DEOHS Bootcamp 2016"
date: "Sept. 16, 2016"
linkcolor: blue
output:
  pdf_document:
    fig_caption: yes
    number_sections: yes
    fig_height: 3
---

\setcounter{tocdepth}{2}
\tableofcontents

# Initial Setup

* Install and load R packages as needed.
```{r pre_load_packages, echo=FALSE, message=FALSE, warning=FALSE}
my_preferred_repo <- 'http://cran.r-project.org'
if (!require("pacman")) {install.packages("pacman", repos = my_preferred_repo)}
pacman::p_load(knitr, plyr, pander, ggplot2)
```
* Disable the code display. The embedded R code will not be echoed in the report.
```{r set_knitr_options, echo=FALSE, message=FALSE, warning=FALSE}
opts_chunk$set(tidy=FALSE, cache=FALSE, echo=FALSE)
```
* Create the `data` folder. It will be created if it does not already exist.
```{r}
# Create the data and image folders, if necessary.
datadir <- "data"
dir.create(file.path(datadir), showWarnings=FALSE, recursive=TRUE)
imagedir <- "images"
dir.create(file.path(imagedir), showWarnings=FALSE, recursive=TRUE)
```
* User-defined functions will be used to help parse the text of survey results.
```{r}
# Define utility functions.

# Count the number of times a regular expression matches items in a vector.
cnt.re <- function(re, v) {
    length(which(grepl(re, v, ignore.case = TRUE)))
}

# Search the text of the responses for specific choices and count the hits.
cnt.choices <- function(choices, re, v, respondents, 
                        colname1='Choice', 
                        colname4='% of Respondents') {
    cnt.df <- ldply(re, function(re) cnt.re(re, v))
    df <- data.frame(Choice=choices, cnt.df)
    names(df) = c(colname1, 'Count')
    df$`% of Counts` <- df$Count / sum(df$Count) * 100
    df$pct_resp <- df$Count / respondents * 100
    names(df)[4] <- colname4
    df <- df[order(df$Count, decreasing = TRUE),]
    row.names(df) <- NULL
    df
}

# Split strings and combine one-word results into a new vector.
parse.responses <- function(v) {
    tolower(gsub('^.*\\s?([A-Z]{3,})\\?? .*$', '\\1', 
         unlist(strsplit(v, '(?:(?:,|\\.|\\band\\b)\\s?)+')), perl=TRUE))
}

# Generate a table of Yes/No response data.
yn.table <- function(content) {
    row.names(content) <- NULL
    pandoc.table(content, justify = c('left', rep('right', 6)), 
             style='multiline', round = 0)
}

# Generate a table of frequency counts for multiple choice data.
count.frequency <- function(x, n, title = "") {
    choice <- unique(unlist(strsplit(x, ',\\s?')))
    cnt.choices(choice, choice, x, n, title)
}

# Calculate the answer rate for a question.
ar <- function(q) round(answer.rate[q] * 100, 0)
```

# Data Import and Cleanup

Data source: [DEOHS Bootcamp - R](http://tinyurl.com/deohs-rboot1), UW Catalyst 
(CSV file)

## Data Import

The steps to read in the data files(s) are as follows:

* Check for existence of data file(s). Abort if they are not found.
```{r}
raw.data.file.name <- 'survey_data.csv'
raw.data.file.path <- file.path(datadir, raw.data.file.name)

questions.file.name <- 'questions_rbootcamp.txt'
questions.file.path <- file.path(datadir, questions.file.name)

for (file in c(raw.data.file.path, questions.file.path)) {
    if (! file.exists(file)) {
        stop(paste("Can't find", file, "!", sep=" "))
    }
}
```
* Read the data file(s), count respondents, and set column names.
```{r}
df <- read.csv(raw.data.file.path, header=FALSE, stringsAsFactors = FALSE, skip = 7)

num_respondents <- nrow(df)

names(df) <- c('start.date', 'finish.date', 'blank', 'q01.dept.role.mc', 
               'q02.r.usage.mc', 'q03.r.script.yn', 'q04.other.lang.yn', 
               'q05.class.goal', 'q06.discussion.group.yn')
```
* Read the questions from a text file.
```{r}
# Make a vector of questions by reading them in from a text file.
questions <- readLines(questions.file.path)
```

The survey had a total of `r nrow(df)` respondents.

## Data Clean-up

The steps to prepare the data for analysis are as follows:

* Replace missing values ('999', 'N/A', 'n/a', 'not applicable', and '') with NA.
```{r}
df[ df == '999' | df == 'N/A' | df == 'n/a' | df == 'not applicable' | df == '' ] <- NA
```
* Calculate the answer rate (AR) for all questions.
```{r}
answer.rate <- sapply(names(df), 
                      function(q) {
                          length(df[complete.cases(df[q]), q]) / num_respondents
                      })
```
* Remove the 'blank' (empty) column included in the original data file.
```{r}
df$blank <- NULL
```


# Additional Processing

## Yes/No Questions

The results for the Yes/No questions are further processed as follows:

* Count the frequency of answers for each of the Yes/No questions.
```{r}
# Count the frequency of answers for each of the Yes/No questions.
yn <- names(df)[grepl( "\\.yn" , names(df))]
yn.df <- data.frame(t(as.data.frame(sapply(df[,yn], table))))
yn.df$id <- row.names(yn.df)
```
* Link the question text to the responses by question ID.
```{r}
# Link the question text to the responses by question ID.
names(questions) <- names(df)
yn.df <- merge(
    data.frame(id = yn, q = questions[yn], stringsAsFactors = FALSE), yn.df)
```
* Clean up row and column names.
```{r}
# Set friendly column names.
names(yn.df) <- c('id', 'Question', 'No', 'Yes')

# Remove row names.
row.names(yn.df) <- NULL
```
* Calculate percentages of response counts per number of responses.
```{r}
# Calculate percentages.
yn.df$Sum <- yn.df$Yes + yn.df$No
yn.df$'% AR' <- yn.df$Sum / num_respondents * 100
yn.df['% Yes'] <- yn.df$Yes / yn.df$Sum * 100
yn.df['% No'] <- yn.df$No / yn.df$Sum * 100
yn.cols <- c('Question', 'Yes', 'No', 'Sum', '% AR', '% Yes', '% No')
```
* Format question column for better display.
```{r}
# Format question column for emphasis Add extra line for better row spacing.
yn.df$Question <- paste('*', yn.df$Question, '*', '\\\n', sep='')
```

# Results

## Primary DEOHS Role

**Question**: *`r questions['q01.dept.role.mc']`*

```{r, results='asis'}
role <- count.frequency(tolower(df$q01.dept.role.mc), num_respondents, 'Role')
```

```{r, results='asis'}
set.caption('Primary DEOHS role frequency counts.')
pandoc.table(role, round = 0, justify = c('left', rep('right', 3)))
```

```{r}
# http://www.cookbook-r.com/Graphs/Bar_and_line_graphs_(ggplot2)/
# No legend, since the information is redundant
ggplot(data=role, aes(x=Role, y=`% of Respondents`, fill=Role)) +
    geom_bar(colour="black", stat="identity") + ggtitle("Primary DEOHS Role") +
    guides(fill=FALSE) + theme_light() + scale_fill_brewer(palette="Set2")
```


## R Usage

**Question**: *`r questions['q02.r.usage.mc']`*

```{r, results='asis'}
usage <- count.frequency(tolower(df$q02.r.usage.mc), num_respondents, 'Usage')
```

```{r, results='asis'}
set.caption('R usage frequency counts.')
pandoc.table(usage, round = 0, justify = c('left', rep('right', 3)))
```

```{r}
# http://www.cookbook-r.com/Graphs/Bar_and_line_graphs_(ggplot2)/
# No legend, since the information is redundant
ggplot(data=usage, aes(x=Usage, y=`% of Respondents`, fill=Usage)) +
    geom_bar(colour="black", stat="identity") + ggtitle("R Usage") +
    guides(fill=FALSE) + theme_light() + 
    scale_fill_brewer(palette="Set2")
```

## Programming Experience

```{r, results='asis'}
content <- yn.df[yn.df$id == 'q03.r.script.yn' | 
                 yn.df$id == 'q04.other.lang.yn', yn.cols]
set.caption('R scripting and other programming languages')
yn.table(content)
```

```{r}
bp <- c('Wrtten R Scripts', 'Used Other Language')
yn.bp <- data.frame(Experience=bp, 'Prevalence.Percent'=yn.df[c(1:2), '% Yes'])
mean.prev <- mean(yn.bp$Prevalence.Percent, round=0)
sd.prev <- sd(yn.bp$Prevalence.Percent)
ggplot(data=yn.bp, aes(x=reorder(Experience, Prevalence.Percent), 
                       y=Prevalence.Percent, fill=Experience)) + 
    geom_bar(stat = "identity", color="black") + coord_flip() + 
    ggtitle("Programming Experience") + theme_light() +
    labs(y = "% Experienced", x = 'Experience') + guides(fill=FALSE) +
    scale_fill_brewer(palette="Set2")
```


## Participant Goals

**Question**: *`r questions['q05.class.goal']`*

These are all of the responses for this question:

```{r, results='asis'}
response <- data.frame(
    Response=df$q05.class.goal[complete.cases(df$q05.class.goal)])
set.caption('Class goals, all responses.')
pandoc.table(response, style='rmarkdown', justify='left')
```
